A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available online \url{\si{http://web.engr.illinois.edu/\string~\si{raichel2}/\si{paint.pdf}}}}

### 2014-11-04 paint-contour-tree.tex
done with first pass

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\def \STOC {}

@@

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(t\alpha(t)+N)$ for balanced trees. 
Moreover, we prove a strong optimality property of our algorithm, showing that $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. Consider the family of instances where our algorithm
requires $C$ comparisons. Any algorithm computing contour trees on this family requires
$\Omega(C)$ comparisons in the worst case.

@@

\begin{theorem} 
\label{thm:main-alg} 
Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

@@

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
specific path decomposition $P(T)$%
\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

%\begin{definition}\label{def:hfdt}
%A \emph{height function decision tree} is an algorithm
%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
%tree over the heights. The running time is simply the number of queries made.
%\end{definition}
%
\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

@@

\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
computing contour trees is expensive, and provide insight into the hardness of particular instances. It is intriguing
that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
Our result is related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

The following is the analogous statement to \Thm{main-lb}, but for join trees.
Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.

\begin{theorem}\label{thm:joinLB}
Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
Any height function decision tree that correctly computes
the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}
\begin{proof}
 If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
 By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
 (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
 Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
 have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

\subsection{Contour Trees}
We first generalize previous terms to the case of contour trees.
In this section $T$ will denote an arbitrary contour tree.  Specifically, every internal vertex in $T$ has degree $3$ 
and the height function defined over the vertex set is such that every internal vertex in either a split or join 
(i.e. at most $2$ neighbors lie above or below it).

We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
Again we define a path decomposition as a partition of the vertices into 
disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
However, for simplicity we now restrict our attention to path decompositions which 
roughly speaking are consistent with the raining procedure described in \Sec{rain} 
(more general decompositions can work, but it is not needed for our purposes).

@@

Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
As described in the previous section, this painting procedure defines a maximum path decomposition with 
and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $\mathcal{P} = P(f)$ be the rain consistent maximum path decomposition of some height function $f$ defined over a complex $\MM$. 
Then there is a family of functions, $\bF$, all defined over a common triangulation in plane, 
such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $\bF$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $\mathcal{P}$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

\begin{lemma}
Let $\mathcal{P}$ and $\bF$ be as defined above.  Then for all $f\in \bF$, $\cost(f) = \sum_{p\in P(f)} |p|\log|p| = \sum_{p\in \mathcal{P}} |p|\log |p|$
\end{lemma}
\begin{proof}
 This lemma follows immediately from \Lem{cost}.  Specifically, the sum $\cost(P)$, for any rain consistent maximum path decomposition $P$, can be broken 
 into the sums $\cost_{\jc}(P_i)$ of some maximum path decompositions $P_1, \dots P_k$.  Now apply \Lem{cost} to each of the $P_i$ separately and take the sum.
\end{proof}

Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following.

\begin{theorem}
Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
Any height function decision tree that correctly computes
the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\def \STOC {}

@@

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$.
This is at most $O(t\log D + N)$, where $D$ is the diameter of the contour tree.
Moreover, it is $O(t\alpha(t)+N)$ for balanced trees, a significant improvement over the previous complexity.

Our algorithm requires numerous ideas: partitioning the contour tree into join and split trees, 
a local growing procedure to iteratively build contour trees, and the use of heavy path 
decompositions for the time complexity analysis. There is a crucial use of a family of binomial heaps to 
maintain priorities, ensuring that any comparison made is between comparable nodes of the contour tree.
We also prove lower bounds showing that the $\sum_{p\in P} |p|\log|p|$ complexity
is inherent to computing contour trees. 
%Consider the family of instances where our algorithm
%requires $C$ comparisons. Any algorithm computing contour trees on this family requires
%$\Omega(C)$ comparisons in the worst case.

@@

\begin{theorem} 
\label{thm:main-alg} 
Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is a deterministic algorithm to compute $T$ whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).

The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
\end{theorem}

@@

We complement \Thm{main-alg} with a lower bound, suggesting that the comparison complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is inherent for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case.
Let a path decomposition $P$ be called \emph{valid} if it is actually output by the algorithm
on some input.
%As referenced in \Thm{main-alg}, for any $f:\MM \mapsto \RR$ our deterministic algorithm defines a 
%specific path decomposition $P(T)$%
\ignore{\FullVer{
\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
\SubmitVer{
\footnote{The specific decomposition used is defined in the full version of the paper.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}.}
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for ant $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

@@

%\begin{definition}\label{def:hfdt}
%A \emph{height function decision tree} is an algorithm
%that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
%tree over the heights. The running time is simply the number of queries made.
%\end{definition}
%
}
\begin{theorem} 
\label{thm:main-lb}
%Fix a sufficiently large numbers $t$ and $C$, 
%where $t$ is the number of critical points and $C$ is a valid cost for an instance with $t$ critical point.
Consider a valid path decomposition $P$. There exists a family $\bF_P$ of terrains ($d=2$) with the following properties.
Any contour tree algorithm makes $\Omega(\sum_{p \in P} |p|\log |p|)$ comparisons in the worst case over $\bF_P$.
Furthermore, for any terrain in $\bF_P$, the algorithm of \Thm{main-alg} makes $O(\sum_{p \in P} |p|\log |p|)$ comparisons.
%
%the set of functions $\bF_C$ that have the same number of critical points and the same (valid) cost $C$.  Any algebraic decision tree that correctly computes
%the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}

In other words, for any choice of $\sum_{p \in P} |p|\log |p|$, we can design a hard family where any algorithm
requires this cost in the worst case, and our algorithm matches this bound.

@@

\paragraph{Beyond worst-case analysis:} Our result provides understanding of the complexity
of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound explain \emph{when}
computing contour trees is expensive, and provide insight into the hardness of particular instances. 
%It is intriguing
%that the cost (as defined earlier) roughly captures the hardness of an instance.
Trees that are short and fast are easy to compute, whereas trees with long paths are hard.
This is tangentially related to the concept of instance optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
of easy to hard instances. From a mathematical perspective, this sort of analysis forces one to look closely
at the relationship between sorting and the contour tree problem.

@@

The following is the equivalent of \Thm{main-lb} for join trees, and immediately follows from
the previous lemmas.
%Call a cost $C$ join tree valid if $C = cost_{\jc}(f)$ for some $f$ defined over a given complex $\MM$.

\begin{theorem}\label{thm:joinLB}
Consider a rooted tree $T$ and an arbitrary path decomposition $P$ of $t$.
There is a family $\bF_P$ of terrains such that any algebraic decision
tree computing the contour tree (on $\bF_P$) requires $\Omega(\sum_{p \in P} |p|\log |p|)$ time.
Furthermore, our algorithm makes $O(\sum_{p \in P} |p|\log |p|)$ on all these instances.
%
%Consider the set of functions $\bF_C$ that all have the same join tree valid cost $C$.  
%Any height function decision tree that correctly computes
%the join tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

\begin{proof}
The proof is a basic entropy argument. Any algebraic decision tree that is correct on all of $\bF_P$
must distinguish all inputs in this family. By Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\Lem{cost} completes the proof.
%
%
% If $C$ is a join tree valid cost then there some $f$ such that $\cost_{\jc}(f)=\sum_{p\in P(f)} |p|\log|p|$.
% By the above lemmas, there is then a set $\bF$ of $\Pi_{p_i\in P} |p_i|!$ functions, such that for any $g\in \bF$, $C= \cost(g)$.  
% (Specifically, each sorted ordering of the relative height values of the saddles on each diamond yields a function in this class.)
% Hence $\bF\subseteq \bF_C$, and so any height function decision tree for the set $\bF_C$ must distinguish between these possibilities, and therefore must 
% have $\Pi_{p_i\in P} |p_i|!$ leaves.  Hence by Stirling's approximation, the average depth in this tree is $\Omega(\sum_{p_i\in P} |p_i|\log|p_i|)$.
\end{proof}

\subsection{Contour Trees}
We first generalize previous terms to the case of contour trees.
In this section $T$ will denote an arbitrary contour tree with every internal vertex of degree $3$.
%and the height function defined over the vertex set is such that every internal vertex in either a split or join 
%(i.e. at most $2$ neighbors lie above or below it).

%We need to generalize the definition of path decompositions, which previously were only defined for rooted trees.
%Again we define a path decomposition as a partition of the vertices into 
%disjoint leaf paths (i.e. each path is monotonic in height and contains a leaf). 
For simplicity we now restrict our attention to path decompositions 
consistent with the raining procedure described in \Sec{rain} 
(more general decompositions can work, but it is not needed for our purposes).

@@

%
%Consider our algorithm on a fixed $(f,\MM)$.  First we run $\rain(x,\MM)$ to partition the input 
%into extreme dominant complexes.  To make this partition uniquely defined lets say that the initial vertex $x$
%that rain is called from is the global maximum in the input (note that given a specific $x$ the partition is uniquely defined).
%Let $\MM_1, \dots, \MM_k$ be the set of produced subcomplexes, and let $f_i$ be the restriction of $f$ to $\MM_i$.
%Now on each of the $(f_i,\MM_i)$ we run the $\paint$ procedure to compute the join tree of $(f_i, \MM_i)$.
%As described in the previous section, this painting procedure defines a maximum path decomposition with 
%and associated cost, $\cost_{\jc}(f_i) = \sum_{p\in P_i} |p|\log |p|$.  
%It is not hard to see that the union of these paths decomposition $P= \cup_{i=1}^k P_i$ defines 
%a rain consistent path decomposition of the contour tree for $(f,\MM)$.  
%In particular for a fixed $\MM$, $f$ defines a specific path decomposition $P=P(f)$, which we call the \emph{rain consistent maximum path decomposition}, 
%and we write $\cost(f) = \sum_{p\in P} |p|\log |p|$.

@@

\begin{lemma}
\label{lem:pathDecomp2}
Let $P$ be a rain consistent maximum path decomposition of some contour tree. 
There is a family of terrains, $\bF$, all with the same triangulation, 
such that the size of $\bF$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two functions in $\bF$ define the same contour tree.
\end{lemma}
\begin{proof}
 As $P$ is rain consistent, the paths can be partitioned into sets $P_1, \dots, P_k$, where $P_i$ is the set of all paths with 
 vertices from a given BFS, as described in \Def{path2}.  Specifically, let $T_i$ be the subtree of $T$ corresponding to $P_i$ and 
 let $r_i$ be the root vertex of this subtree.  
 Note that the $P_i$ sets naturally define a tree where $P_i$ is the parent of $P_j$ if $r_i$ (i.e. the root of $T_i$) is adjacent to a vertex in $P_j$.

@@

\begin{lemma}
For all $\MM \in \bF_P$, the number of heap operations is $\Theta(\sum_{p\in P} |p|\log |p|)$
\end{lemma}
\begin{proof}
 This lemma follows immediately from \Lem{cost}.  The heap operations can be partitioned into the
 operations performed in each $P_i$.
 Apply \Lem{cost} to each of the $P_i$ separately and take the sum.
\end{proof}

\Thm{main-lb} follows immediately from an entropy argument, analogous to \Thm{joinLB}.
%
%
%Call a cost $C$ valid if $C = cost(f)$ for some $f$ defined over a given complex $\MM$.
%For the same reason that \Thm{joinLB} followed from \Lem{pathDecomp} and \Lem{cost}, we now have the following. 
%
%\begin{theorem}
%Consider the set of functions $\bF_C$ that all have the same valid cost $C$.  
%Any height function decision tree that correctly computes
%the contour tree on all of $\bF_C$, has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
%\end{theorem}
